---
title: "P8105 Homework 2"
author: "Meher Walia"
date: "2026-10-07"
output: github_document
---

```{r setup, message = FALSE}
library(tidyverse)
library(readxl)
```
  
## Problem 1

```{r}
nyc_transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor::clean_names() |> 
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada) |> 
  mutate(
    across(route1:route11, as.character),
    entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
```

These data describe every entrance and exit to the NYC subway system, with
one row per entrance rather than per station. The variables I retained
identify the station (`line`, `station_name`, and the station's latitude and
longitude), the routes it serves (`route1` through `route11`), and
characteristics of the entrance itself: whether it allows `entry`, whether
there is a `vending` machine, the `entrance_type` (stair, elevator, escalator,
door, and so on), and whether the station is `ada` compliant. My cleaning
steps were to read in the CSV, convert the variable names to lower snake case
with `janitor::clean_names()`, select the variables listed above, coerce the
route columns to character so they share a type, and convert `entry` from the
character values "YES" and "NO" to a logical variable using `case_match()`.
The resulting dataset has `r nrow(nyc_transit_df)` rows and
`r ncol(nyc_transit_df)` columns.

These data are not tidy. The eleven route columns hold the same kind of
information spread across many columns, so a single variable (the route served)
is stored in eleven places rather than one. Most of these columns are also
empty for most entrances, since few stations serve more than two or three
routes.

```{r}
nyc_transit_df |> 
  distinct(line, station_name) |> 
  nrow()

nyc_transit_df |> 
  filter(ada) |> 
  distinct(line, station_name) |> 
  nrow()

nyc_transit_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

There are 465 distinct stations, identified by the combination of station name
and line. Of these, 84 are ADA compliant. Among entrances without a vending
machine, about 37.7% allow entry.

```{r}
nyc_transit_tidy_df = 
  nyc_transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |> 
  drop_na(route_name)

nyc_transit_tidy_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

nyc_transit_tidy_df |> 
  filter(route_name == "A", ada) |> 
  distinct(line, station_name) |> 
  nrow()
```

After reformatting so that route number and route name are separate variables,
60 distinct stations serve the A train. Of those, 17 are ADA compliant.
  
  ## Problem 2

```{r}
clean_trash = function(df, wheel_label) {
  df |> 
    janitor::clean_names() |> 
    filter(!is.na(dumpster)) |> 
    mutate(
      across(any_of(c("dumpster", "year", "weight_tons", "volume_cubic_yards",
                      "plastic_bottles", "polystyrene", "cigarette_butts",
                      "glass_bottles", "plastic_bags", "wrappers",
                      "sports_balls", "homes_powered")),
             ~ suppressWarnings(as.numeric(.x))),
      trash_wheel = wheel_label)
}

trash_path = "./data/202610 Trash Wheel Collection Data.xlsx"

mr_trash_df = 
  read_excel(trash_path, sheet = "Mr. Trash Wheel", range = cell_cols("A:N")) |> 
  clean_trash("Mr. Trash Wheel") |> 
  mutate(sports_balls = as.integer(round(sports_balls)))

professor_trash_df = 
  read_excel(trash_path, sheet = "Professor Trash Wheel", range = cell_cols("A:N")) |> 
  clean_trash("Professor Trash Wheel")

gwynnda_trash_df = 
  read_excel(trash_path, sheet = "Gwynnda the Good Wheel of the W", range = cell_cols("A:N")) |> 
  clean_trash("Gwynnda")

trash_wheel_df = 
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_trash_df) |> 
  select(-starts_with("x")) |> 
  relocate(trash_wheel)
```

These data record each dumpster of trash collected by three water wheels in
the Baltimore harbor. Each row is one dumpster, identified by the
`trash_wheel` that collected it, its `dumpster` number, and the `month`,
`year`, and `date` of collection. The remaining variables measure what was
collected: the total `weight_tons` and `volume_cubic_yards`, counts of
specific items such as `plastic_bottles`, `cigarette_butts`, `wrappers`, and
`sports_balls`, and the number of `homes_powered` by the resulting energy.
For each sheet I read only columns A through N to exclude the note columns,
dropped rows without a dumpster number, coerced the count variables to
numeric, and added a variable identifying the wheel; for Mr. Trash Wheel I
also rounded sports balls to the nearest integer. Combining the three sheets
gives `r nrow(trash_wheel_df)` observations across
`r ncol(trash_wheel_df)` variables. Not every wheel tracks every item, so
columns like `sports_balls` are missing for the wheels that do not record them.

```{r}
trash_wheel_df |> 
  filter(trash_wheel == "Professor Trash Wheel") |> 
  pull(weight_tons) |> 
  sum(na.rm = TRUE)

trash_wheel_df |> 
  filter(trash_wheel == "Gwynnda", year == 2022, month == "June") |> 
  pull(cigarette_butts) |> 
  sum(na.rm = TRUE)
```

Across the available data, Professor Trash Wheel collected a total of 294.96
tons of trash. Gwynnda collected 18,120 cigarette butts in June of 2022.