---
title: "p8105_hw2_mw3952.Rmd"
author: "Meher Walia"
date: "2026-10-07"
output: github_document
---

```{r setup, message = FALSE}
library(tidyverse)
```

## Problem 1

```{r}
nyc_transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor::clean_names() |> 
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))

nyc_transit_df
```

list.files("./data")

# how many distinct stations
nyc_transit_df |> 
  distinct(line, station_name) |> 
  nrow()

# how many are ADA compliant
nyc_transit_df |> 
  filter(ada) |> 
  distinct(line, station_name) |> 
  nrow()

# proportion of non-vending entrances that allow entry
nyc_transit_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
  
  nyc_transit_tidy_df = 
  nyc_transit_df |> 
  mutate(across(route1:route11, as.character)) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |> 
  drop_na(route_name)

nyc_transit_tidy_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

nyc_transit_tidy_df |> 
  filter(route_name == "A", ada) |> 
  distinct(line, station_name) |> 
  nrow()
  
  ## Problem 1
## Problem 1

```{r}
nyc_transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor::clean_names() |> 
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada) |> 
  mutate(
    across(route1:route11, as.character),
    entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
```

These data describe every entrance and exit to the NYC subway system, with
one row per entrance rather than per station. The variables I retained
identify the station (`line`, `station_name`, and the station's latitude and
longitude), the routes it serves (`route1` through `route11`), and
characteristics of the entrance itself: whether it allows `entry`, whether
there is a `vending` machine, the `entrance_type` (stair, elevator, escalator,
door, and so on), and whether the station is `ada` compliant. My cleaning
steps were to read in the CSV, convert the variable names to lower snake case
with `janitor::clean_names()`, select the variables listed above, coerce the
route columns to character so they share a type, and convert `entry` from the
character values "YES" and "NO" to a logical variable using `case_match()`.
The resulting dataset has `r nrow(nyc_transit_df)` rows and
`r ncol(nyc_transit_df)` columns.

These data are not tidy. The eleven route columns hold the same kind of
information spread across many columns, so a single variable (the route served)
is stored in eleven places rather than one. Most of these columns are also
empty for most entrances, since few stations serve more than two or three
routes.

```{r}
nyc_transit_df |> 
  distinct(line, station_name) |> 
  nrow()

nyc_transit_df |> 
  filter(ada) |> 
  distinct(line, station_name) |> 
  nrow()

nyc_transit_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

There are 465 distinct stations, identified by the combination of station name
and line. Of these, 84 are ADA compliant. Among entrances without a vending
machine, about 37.7% allow entry.

```{r}
nyc_transit_tidy_df = 
  nyc_transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |> 
  drop_na(route_name)

nyc_transit_tidy_df |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

nyc_transit_tidy_df |> 
  filter(route_name == "A", ada) |> 
  distinct(line, station_name) |> 
  nrow()
```

After reformatting so that route number and route name are separate variables,
60 distinct stations serve the A train. Of those, 17 are ADA compliant.

